---
title: "Wine Analysis Using R"
author: "Mason Krug"
date: "5/1/2021"
output: html_document
---

# Introduction to the Data

List of packages used in this project:
```{r Libraries, message=FALSE, warning=FALSE, include=TRUE}

# This will simply print the libraries used

library(ggplot2)
library(gridExtra)
library(ggcorrplot)
```

```{r Data Preparation, message=FALSE, warning=FALSE, include=FALSE}

# This imports the CSV file for manipulation from R

wine.red <- read.csv('wineQualityReds.csv')
```

> **Data Introduction**: The data set that will be analyzed consists of 1599 entries of information about red wine. Variables that will be analyzed include quality, alcohol content, pH levels, residual sugar content, and more.

# Univariate Plots Section

> **Structure of the Data**: Here we can see the structure that the CSV file takes. It consists of the following variables. Quality of the wine is rated from 1 - 10 and alcohol represents alcohol level as a percentage. This data set consists of 13 variables to explore.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# This details the type in each variable (string/int) and gives a small sample from each column

str(wine.red)
```

> **Data Summary**: Below is a summary of the data from the CSV file. Here we can understand a bit more about each variable and the spread of each variable.

```{r Data summary, echo=FALSE, message=FALSE, warning=FALSE}

# This shows the structure of the data in the CSV file.IE: mean, median, and more of each variable

summary(wine.red)
```

<!-- Here we will create a categorical relationship of the wine quality. A description of the quality is sometimes easier to interpret than a number representing the quality. -->

```{r Create Quality Index, echo=FALSE, message=FALSE, warning=FALSE}
wine.red$quality.cat <- NA
wine.red$quality.cat <- ifelse(wine.red$quality >= 9, 'Superb', NA)
wine.red$quality.cat <- ifelse(wine.red$quality == 8, 'Exceptional', NA)
wine.red$quality.cat <- ifelse(wine.red$quality == 7, 'Great', NA)
wine.red$quality.cat <- ifelse(wine.red$quality == 6, 'Good', NA)
wine.red$quality.cat <- ifelse(wine.red$quality == 5, 'Acceptable', NA)
wine.red$quality.cat <- ifelse(wine.red$quality == 4, 'Mediocre', NA)
wine.red$quality.cat <- ifelse(wine.red$quality == 3, 'Bad', NA)
wine.red$quality.cat <- ifelse(wine.red$quality == 2, 'Bad', NA)
wine.red$quality.cat <- ifelse(wine.red$quality == 1, 'Nasty', NA)
wine.red$quality.cat <- factor(wine.red$quality.cat, levels = c("Superb", "Exceptional", "Great", "Good", "Acceptable", "Bad", "Nasty"))
```

# Univariate Analysis

> **Histogram of Wine Quality**: Here we can see that a majority of the wine has a quality of 5 or 6 with a handfull of wines labeled as 7.

```{r Histogram of Wine Quality, echo=FALSE, message=FALSE, warning=FALSE}

# This creates a histogram showing the count of wines sorted by quality

ggplot(aes(x = quality), data = wine.red) +
  geom_histogram(binwidth = 1, color = 'black', fill = '#660613') +
  scale_x_continuous(breaks = seq(1,8,1)) +
  geom_text(aes(label = ..count..), stat = "count", vjust = -0.2, colour = "black") +
  xlab('Quality of Wines') + 
  ylab('Number of Wines') +
  ggtitle('Histogram Showcasing Wine Quality')
  
table(wine.red$quality)
```

> **Further Exploration**: Here I want to graphically represent a handful of vaiables. I want to see if there is any approximation to a normal distribution.

```{r Univariate Analysis Graphs 1, echo=FALSE, message=FALSE, warning=FALSE}

# The below will make 10 charts. Each chart explores 1 variable. They will later be displayed next to one another

h1 <- ggplot(aes(x = fixed.acidity), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  scale_x_continuous(limits = c(4, 16)) +
  scale_y_continuous(trans = 'log10') +
  ggtitle('Histogram of Fixed Acidity') + 
  labs(y  = "Count of Wines") +
  labs(x = "Tartaric Acid in g / dm^3")

h2 <- ggplot(aes(x = volatile.acidity), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  scale_x_continuous(limits = c(0, 2)) +
  scale_y_continuous(trans = 'log10') +
  ggtitle('Histogram of Volatile Acidity') + 
  labs(y  = "Count of Wines") +
  labs(x = "Acetic Acid in g / dm^3")

h3 <- ggplot(aes(x = citric.acid), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(trans = 'log10') +
  ggtitle('Histogram of Citric Acid') + 
  labs(y  = "Count of Wines") +
  labs(x = "Citric Acid in g / dm^3")

h4 <- ggplot(aes(x = residual.sugar), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  scale_x_continuous(limits = c(0, 16)) +
  scale_y_continuous(trans = 'log10') +
  ggtitle('Histogram of Residual Sugar') + 
  labs(y  = "Count of Wines") +
  labs(x = "Residual Sugar in g / dm^3")

h5 <- ggplot(aes(x = chlorides), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(trans = 'log10') +
  ggtitle('Histogram of Chlorides') + 
  labs(y  = "Count of Wines") +
  labs(x = "Sodium Chloride in g / dm^3")

h6 <- ggplot(aes(x = free.sulfur.dioxide), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  scale_x_continuous(limits = c(1, 72)) +
  scale_y_continuous(trans = 'log10') +
  ggtitle('Histogram of Free Sulfur Dioxide') + 
  labs(y  = "Count of Wines") +
  labs(x = "Free Sulfur Dioxide in mg / dm^3")

h7 <- ggplot(aes(x = total.sulfur.dioxide), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  scale_x_continuous(limits = c(6, 289)) +
  scale_y_continuous(trans = 'log10') +
  ggtitle('Histogram of Total Sulfur Dioxide') + 
  labs(y  = "Count of Wines") +
  labs(x = "Total Sulfur Dioxide in mg / dm^3")


h8 <- ggplot(aes(x = pH), data = wine.red) +
  scale_x_continuous(limits = c(2, 4)) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  scale_y_continuous(trans = 'log10') +
  ggtitle('Histogram of pH') + 
  labs(y  = "Count of Wines") +
  labs(x = "pH")

h9 <- ggplot(aes(x = sulphates), data = wine.red) +
  scale_x_continuous(limits = c(0, 2)) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  scale_y_continuous(trans = 'log10') +
  ggtitle('Histogram of Sulphates') + 
  labs(y  = "Count of Wines") +
  labs(x = "Potassium Sulphate in g / dm^3")

h10 <- ggplot(aes(x = alcohol), data = wine.red) +
  scale_x_continuous(limits = c(8, 15)) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  ggtitle('Histogram of Alcohol') + 
  labs(y  = "Count of Wines") +
  labs(x = "% Alcohol By Volume")

# Displays the above charts in a table-like format
grid.arrange(h1, h2, ncol = 2)
grid.arrange(h3, h4, ncol = 2)
grid.arrange(h5, h6, ncol = 2)
grid.arrange(h7, h8, ncol = 2)
grid.arrange(h9, h10, ncol = 2)
  
```


> **Logarithmic Exploration**: Many of the above histograms are close to a normal distribution. Let's apply a logarithmic function to the above histograms to better approximate a normal distribution.


```{r Univariate Analysis Graphs 2, echo=FALSE, message=FALSE, warning=FALSE}


# The below will make 10 charts. Each chart explores 1 variable. They will later be displayed next to one another

l1 <- ggplot(aes(x = log(fixed.acidity)), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  ggtitle('Log 10 of Histogram of Fixed Acidity') + 
  labs(y  = "Count of Wines") +
  labs(x = "Logarithmatic Tartaric Acid in g / dm^3")

l2 <- ggplot(aes(x = log(volatile.acidity)), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  scale_x_continuous(limits = c(0, 2)) +
  ggtitle('Log 10 of Histogram of Volatile Acidity') + 
  labs(y  = "Count of Wines") +
  labs(x = "Log10 of Acetic Acid in g / dm^3")

l3 <- ggplot(aes(x = log(citric.acid)), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  ggtitle('Log 10 of Histogram of Citric Acid') + 
  labs(y  = "Count of Wines") +
  labs(x = "Log10 of Citric Acid in g / dm^3")

l4 <- ggplot(aes(x = log(residual.sugar)), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  scale_x_continuous(limits = c(0, 16)) +
  ggtitle('Log 10 of Histogram of Residual Sugar') + 
  labs(y  = "Count of Wines") +
  labs(x = "Log10 of Residual Sugar in g / dm^3")

l5 <- ggplot(aes(x = log(chlorides)), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  ggtitle('Log 10 of Histogram of Chlorides') + 
  labs(y  = "Count of Wines") +
  labs(x = "Log10 of Sodium Chloride in g / dm^3")

l6 <- ggplot(aes(x = log(free.sulfur.dioxide)), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  ggtitle('Log 10 of Histogram of Free Sulfur Dioxide') + 
  labs(y  = "Count of Wines") +
  labs(x = "Log10 of Free Sulfur Dioxide in mg / dm^3")

l7 <- ggplot(aes(x = log(total.sulfur.dioxide)), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  ggtitle('Log 10 of Histogram of Total Sulfur Dioxide') + 
  labs(y  = "Count of Wines") +
  labs(x = "Log10 of Total Sulphur Dioxide in mg / dm^3")

l8 <- ggplot(aes(x = log(pH)), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  ggtitle('Log 10 of Histogram of pH') + 
  labs(y  = "Count of Wines") +
  labs(x = "Log10 of pH")

l9 <- ggplot(aes(x = log(sulphates)), data = wine.red) +
  scale_x_continuous(limits = c(0, 2)) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  ggtitle('Log 10 of Histogram of Sulphates') + 
  labs(y  = "Count of Wines") +
  labs(x = "Log10 of Potassium Sulphate in g / dm^3")

l10 <- ggplot(aes(x = log(alcohol)), data = wine.red) +
  geom_histogram(color = 'black', fill = '#660613', bins = 30) +
  ggtitle('Log 10 of Histogram of Alcohol') + 
  labs(y  = "Count of Wines") +
  labs(x = "Log10 of % Alcohol By Volume")

# Displays the above charts in a table-like format
grid.arrange(l1, l2, ncol = 2)
grid.arrange(l3, l4, ncol = 2)
grid.arrange(l5, l6, ncol = 2)
grid.arrange(l7, l8, ncol = 2)
grid.arrange(l9, l10, ncol = 2)
  
```

> Above we can see that some variables are much better represented using Log10. Fixed acidity, volatile acidity, sulphates, and alcohol are now more normal. The other graphs approximate normal as well however outliers greatly skew the results.

### What is the structure of your dataset?

> The data set is composed of 1599 entries expanding 13 variables.

### What is/are the main feature(s) of interest in your dataset?

> I am interested in the quality of wine and alcohol content as they relate to other variables in the dataset.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

> I believe sugar content will play a large role in perceived quality and I believe it will be inversely related to alcohol content. Yeast consumes sugar to produce alcohol so it stands to reason that the more alcohol content a wine has, the less sugar it will have.

### Did you create any new variables from existing variables in the dataset?

> I created a categorical variable from the quality of wine, where the 1-10 rating is labeled as follows:

* 9 and 10 are Superb
* 8 is Exceptional
* 7 is Great
* 6 is Good
* 5 is Acceptable
* 4 is Mediocre
* 3 and 2 are Bad
* 1 is Nasty

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

> I wanted to compare the count of wines against a single variable to see if there was a relationship to a normal distribution. Fixed Acidity, Volatile Acidity, pH, Sulphate Levels, and Alcohol content were roughly normal in distribution. Sulfur levels and Free Sulphur Dioxides approximated the right half of a normal distribution however they had a handful of outliers. Residual Sugar levels approximated a normal distribution however also showed a notable number of outliers on the right-hand side.

# Bivariate Analysis

### Plot One

```{r Correlation Investigation 1, echo=FALSE, message=FALSE, warning=FALSE}

# This function finds the correlation columns 2 - 13 of the CSV, finds the correlation between each variable, rounds it off, and then plots it in a correlation graph.

corr <- round(cor(wine.red[, 2:13]), 3)
ggcorrplot(corr, hc.order = TRUE,
           outline.col = "gray",
           lab = TRUE,
           tl.cex = 7,
           digits = 2,
           tl.srt = 90,
           lab_size = 3,
           p.mat = NULL,
           insig = c("pch", "blank"),
           pch = 1,
           pch.col = "black",
           pch.cex =1)
```

### Plot Two

```{r Correlation Investigation 2, echo=FALSE, message=FALSE, warning=FALSE}

# This charts alcohol content against quality

wine.red$quality_factor <- factor(wine.red$quality)
ggplot(data = wine.red, aes(y = alcohol, x = quality, color = quality_factor)) +
  geom_jitter(alpha = 1/4) + 
  geom_smooth(method = "lm")

```




### Plot Three

```{r Correlation Investigation 3, echo=FALSE, message=FALSE, warning=FALSE}

#Creates a box plot comparing alcohol content and quality

boxplot1 <- ggplot(data = wine.red, aes(y = alcohol, x = quality_factor)) +
  geom_boxplot()
boxplot1

```


### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

> pH and Fixed Acidity as well as pH and Volatile Acidity Acid levels had the strongest relationship of __-.060__ and __-.055__, respectively.

#### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

> The density of the wine was not something I would expect to have a strong relationship with other variables. Upon further thinking on the topic, it makes sense that a higher residual sugar level would correlate strongly with density (__0.36__). Additionally, alcohol level and density are also strongly correlated (__0.50__). It makes sense that the more solutes you have in solution, the more dense the solution would be. 

### What was the strongest relationship you found?

> The strongest relationship observed was betwenn fixed acidity and citric acid (r = ___0.67___).

# Multivariate Analysis

### Plot One

```{r Multivariate Plots 1, echo=FALSE, message=FALSE, warning=FALSE}
  scale_color_gradient(high = 'green', low = 'red')
ggplot(wine.red, aes(x = alcohol, y = residual.sugar, color = quality)) +
  geom_jitter() +
  scale_color_gradient(high = 'green', low = 'red')
```

### Description One

> My initial assumption about alcohol content and residual sugar level appears to be incorrect. There is a very weak relationship between the amount of alcohol in these wine samples and the residual sugar level. Residual sugar level is most strongly correlated with density and the strongest relationship that density has with another variable is fixed accidity. Let's explore that below.

### Plot Two

```{r Multivariate Plots 2, echo=FALSE, message=FALSE, warning=FALSE}
  scale_color_gradient(high = 'green', low = 'red')
ggplot(wine.red, aes(x = density, y = fixed.acidity, color = alcohol)) +
  geom_jitter() +
  scale_color_gradient(high = 'green', low = 'red')
```

### Description Two

> Here we can see that strong relationships explained above. Density correlates with fixed acidity at __r = 0.67__ and fixed alcohol content correlates strongest with density at __r = -0.50__. Below we will visually investigate the variable quality.

### Plot Three

```{r Multivariate Plots 3, echo=FALSE, message=FALSE, warning=FALSE}
  scale_color_gradient(high = 'green', low = 'red')
ggplot(wine.red, aes(x = residual.sugar, y = alcohol, color = quality)) +
  geom_jitter() +
  scale_color_gradient(high = 'green', low = 'red')
```

### Description Three

> Residual sugar has a notable amount of outliers. Barring the outliers, the quality of the wine increases with alcohol content. As alcohol content increases, the spread of residual sugar tends to decrease however outliers strongly influence that correlation coefficient.

### Plot Four

```{r Multivariate Plots 4, echo=FALSE, message=FALSE, warning=FALSE}
  
#Makes a scatter plot comparing pH, density, and quality

  scale_color_gradient(high = 'green', low = 'red')
ggplot(wine.red, aes(x = density, y = pH, color = quality)) +
  geom_jitter() +
  scale_color_gradient(high = 'green', low = 'red')
```

### Description Four

> Densiy and pH have a coeficient of __r = 0.34__. This is a weak to medium strength relationship. When visually graphed, the relationship is apparent and appears to be influenced strongly by outliers. Moreover, it appears that quality has a noticeable relationship, although weak, where higher quality wines are more dense. This is expected because, as explored above, higher quality wines tend to have more alcohol. More alcohol in a wine increases its density.

# Summary of Plots

```{r Histogram of Wine Quality Review, echo=FALSE, message=FALSE, warning=FALSE}

# This creates a histogram showing the count of wines sorted by quality

ggplot(aes(x = quality), data = wine.red) +
  geom_histogram(binwidth = 1, color = 'black', fill = '#660613') +
  scale_x_continuous(breaks = seq(1,8,1)) +
  geom_text(aes(label = ..count..), stat = "count", vjust = -0.2, colour = "black") +
  xlab('Quality of Wines') + 
  ylab('Number of Wines') +
  ggtitle('Histogram Showcasing Wine Quality')
  
table(wine.red$quality)
```

> I find it interesting to note how closely this sample approximates a normal distribution. I suspect that if wine ratings had a smaller interval value (ie rating in increments of 0.5) that this would more approximate a normal distribution.

```{r Multivariate Plots 2 Review, echo=FALSE, message=FALSE, warning=FALSE}
  scale_color_gradient(high = 'green', low = 'red')
ggplot(wine.red, aes(x = density, y = fixed.acidity, color = alcohol)) +
  geom_jitter() +
  scale_color_gradient(high = 'green', low = 'red')
```

> This plot from the multivariate section is especially interesting. One item to note that alcohol in solution is acidic; it allows for the free exchange of potential hydrogen (ph) when dissolved in water. One thing to note is that we do see that relationship. As alcohol content increases, the pH level decreases (becomes more acidic). Acidity level in wine is influenced by acohol content, tartaric acid levels, citric acid levels, and possibly more acids in the wine.

```{r Correlation Investigation 1 Review, echo=FALSE, message=FALSE, warning=FALSE}

# This function finds the correlation columns 2 - 13 of the CSV, finds the correlation between each variable, rounds it off, and then plots it in a correlation graph.

corr <- round(cor(wine.red[, 2:13]), 3)
ggcorrplot(corr, hc.order = TRUE,
           outline.col = "gray",
           lab = TRUE,
           tl.cex = 7,
           digits = 2,
           tl.srt = 90,
           lab_size = 3,
           p.mat = NULL,
           insig = c("pch", "blank"),
           pch = 1,
           pch.col = "black",
           pch.cex =1)
```

> I would like to touch on this correlation chart once more. It makes me wonder why certain variables about these wines were chosen to be measured. My guess is that they are important for the taste of wines and contribute to qualities of a wine. However, many variables have a weak relationship with each other which makes me think that, from an analytical perspective, they do not serve much pupose. I suspect that they play a large role in wine taste and so a large correlation might not be absolutely necesary.


# Reflection

> My initial assumption of an inverse relationship between residual sugar and alcohol content appears to be incorrect. While it is true that yeast consumes sugar to create alcohol, it appears that there are more factors at play. I initially struggled with what variables to explore. I am not a wine connoisseur and had no idea that sulfur content, chloride levels, and citric acid content were traits of wine that were notable, let alone even measures. For future analysis, I think it would be interesting to find relationships between quality and year of production. I would be curious to know, year after year, what traits are desired in wines that are high quality. Also, it is unclear if all variables are relevant. What variables are important to track? Which variables would a business like to focus on? These are questions that interest me.